---
title: "Assignment week 8"
author: "Adrian Molière"
date: "2023-05-02"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(epiwraps)
  library(motifmatchr)
  library(MotifDb)
  library(universalmotif)
  library(sechm)
  library(BiocParallel)
  library(ggplot2)
  library(chromVAR) # for motif accessibility estimation
  library(SummarizedExperiment) # data structure
  library(limma) # for statistical analysis

})
# to control multithreading, unix users can use:
# register(MulticoreParam(4))
# for windows users, rather one of the following:
register(SerialParam())
# register(SnowParam(2))
```


## Download the data
```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/mouse_mm38_hippocampus.peakCounts.SE.rds", "ATAC_seq_mouse_hippo.peakCounts.SE.rds")
```


```{r}
SE <- readRDS("ATAC_seq_mouse_hippo.peakCounts.SE.rds")
```

## Get the database of mouse motifs

```{r}
motifs <- query(MotifDb, c("Mmusculus", "HOCOMOCOv10"))
motifs <- do.call(TFBSTools::PWMatrixList, setNames(
           universalmotif::convert_motifs(motifs, class="TFBSTools-PWMatrix"),
           mcols(motifs)$geneSymbol))

names(motifs) = make.names(names(motifs), unique=TRUE)
```
I made the names unique since some motifs were present multiple times.

```{r}
genome <- BiocManager::install("BSgenome.Mmusculus.UCSC.mm10")
```


##Adding factors for condition and sex

```{r}
SE$condition <- as.factor(rep(1:2, each = 6))
levels(SE$condition) <- c("CTRL", "FSS")

SE$sex <- as.factor(rep(1:2, each = 3))
levels(SE$sex) <- c("FEMALE", "MALE")
```

##Adding GC bias


```{r}
seqlevelsStyle(SE) = "UCSC"
SE_gc <- suppressWarnings(chromVAR::addGCBias(SE, genome=genome))

```

#compute deviations and variability

```{r}
moi <- suppressWarnings(motifmatchr::matchMotifs(motifs, subject=SE_gc, genome=genome))


dev <- chromVAR::computeDeviations(object = SE_gc, annotations=moi,
                                   background_peaks=getBackgroundPeaks(SE_gc, niterations=1000))
dev
colData(dev)

variability <- computeVariability(dev)
plotVariability(variability)
```


```{r}
head(assays(dev)$z)
assays(dev)$norm <- scale(assays(dev)$z)
dev
```

#model

```{r}

mm <- model.matrix(~sex+condition, data=as.data.frame(colData(dev)))
mm

```

```{r}
fit <- eBayes(lmFit(assays(dev)$norm, mm))

res <- as.data.frame(topTable(fit, coef="conditionFSS", number = Inf))
head(res)

res$TF <- row.names(res)
ggplot(res, aes(logFC, -log10(adj.P.Val), label=TF)) + geom_text() 
```

The transcription factors with the greatest change in accessibility after FSS are ANDR (Androgen receptor) and GCR (Glucocorticoid receptor). 
The androgen receptor gets activated by binding testosterone, which is released during exercise. Thus it makes sense that the receptor is more active during forced swimming.
The glucocorticoid response plays a crucial role in the "fight or flight" response and thus also makes sense in the context of forced swimming stress. 


```{r}
sechm(dev, features = head(row.names(res)), assayName="norm", top_annotation = c("sex", "condition", "depth"))
```

Looking at the heatmap, we see that sex has very little effect on the expression of GCR and ANDR. 


```{r}
res_sex <- as.data.frame(topTable(fit, coef="sexMALE", number = Inf))
head(res_sex)

res_sex$TF <- row.names(res_sex)
ggplot(res_sex, aes(logFC, -log10(adj.P.Val), label=TF)) + geom_text() 
```
Comparing the signals of the difference sexes reveals that most of the differentially active TFs are less active in male mice. The strongest hits are TYY1 (Yin and Yang 1 protein) and TEAD1 (transcriptional enhancer factor 1). YY1 is important for X chromosome inactivation, thus it makes sense that it is predominantly active in females. I could not find any information why TEAD1 is more active in females than in males. 


```{r}
sechm(dev, features = head(row.names(res_sex)), assayName="norm", top_annotation = c("sex", "condition", "depth"))
```

Looking at the heatmap, we see that at least for the strongest hits the condition has no effect on the activity of the sex-specific transcription factors. 







